---
title: "Association Analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

```{r}
library(dplyr)
library(data.table)
library(ggplot2)
library(parallel)
library(glue)
library(tidyr)
```

## Background

-   get PCs
-   get depression status
-   identify individuals who did not have depression at baseline
-   Read in PRS scores for each ancestry, and protein levels
-   Regress protein \~ PRS for each protein and ancestry, including covariates
-   Lookup agreement of assocs across ancestries
-   Regression in controls only (i.e. excluding individuals who had first episode before enrollment date)


## Notes

- AFR, SAS, EAS have low power, so if you find a protein assoc in those populations, it's quite interesting to see if they don't associate in EUR because EUR have so much power
- For each of AFR, SAS and EAS, select their significant proteins, and perform heterogeneity (like in rcovs_sig2), to see if there's any evidence that these are ancestry specific
- For EUR, we're finding almost half of tested proteins actually have an association. If we don't see heterogeneity for the EUR associations, it suggests that there's a widespread effect in all ancestries, it's not just EUR

- Manhattan plot or something that shows Q for each significant protein
- Look at heterogeneity of SAS, AFR, EAS discovered proteins - this will reduce multiple testing
- For anything that shows heterogeneity, plot the effects in all ancestries to illustrate which ones are showing ancestry-specific effects (e.g. acta2 having no effect in EUR)
- expected-vs-observed analysis, where we ask if the degree of replication of the 732 EUR proteins, is high in the other populations, after accounting for differences in statistical power. Initial analysis looks like it has quite hight replication after accounting for power


## Read in data

```{r}
#load(file.path("~/Downloads", "ukb_prot_mdd.rdata"))
load(file.path("C:/Users/HP OMEN GAMING/Desktop/mr-uganda-hackathon-2023-6/data", "ukb_prot_mdd.rdata"))
head(prot4)
head(dat)
str(prot4)
str(dat)
pops <- c("AFR", "EUR", "EAS", "SAS")
```

## Perform associations

```{r}
fast_assoc <- function(y, x) {
	index <- is.finite(y) & is.finite(x)
	n <- sum(index)
	y <- y[index]
	x <- x[index]
	#vx <- var(x)
	vy <- var(y)
	bhat <- stats::cov(y, x) / vy
	ahat <- mean(x) - bhat * mean(y)
	# fitted <- ahat + x * bhat
	# residuals <- y - fitted
	# SSR <- sum((residuals - mean(residuals))^2)
	# SSF <- sum((fitted - mean(fitted))^2)

	rsq <- (bhat * vy)^2 / (vy * var(x))
	fval <- rsq * (n-2) / (1-rsq)
	tval <- sqrt(fval)
	se <- abs(bhat / tval)

	# Fval <- (SSF) / (SSR/(n-2))
	# pval <- pf(Fval, 1, n-2, lowe=F)
	p <- stats::pf(fval, 1, n-2, lower.tail=FALSE)
	return(list(
		ahat=ahat, bhat=bhat, se=se, fval=fval, pval=p, n=n
	))
}

phewas <- function(y, g, logistic=FALSE) {
	out <- matrix(0, ncol(g), 6)
	if(logistic)
	{
		stopifnot(all(y %in% c(0,1)))
		for(i in 1:ncol(g))
		{
			o <- logistic_assoc(y, g[,i])
			out[i, ] <- unlist(o)
		}
	} else {
		for(i in 1:ncol(g))
		{
			o <- fast_assoc(y, g[,i])
			out[i, ] <- unlist(o)
		}
	}

	out <- dplyr::as_tibble(out, .name_repair="minimal")
	names(out) <- names(o)
	out$snp <- 1:ncol(g)
	return(out)
}
```

Perform simple analysis - no covariates etc

```{r}
rsimple <- lapply(pops, \(p) {
  x <- prot4[[p]]
  o <- phewas(y=unlist(x$normscore), g=as.matrix(as.matrix(x[,4:ncol(x)])))
  o$pop=p
  o$prot <- colnames(x[4:ncol(x)])
  o$fdr <- p.adjust(o$pval, "fdr")
  return(o)
}) %>% bind_rows()

rsimple %>% group_by(pop) %>%
  summarise(n=sum(fdr < 0.05))
```

Try comparing across ancestries

```{r}
rw <- spread(rsimple %>% select(pop, bhat, prot), key=pop, value=bhat)
head(rw)
```

```{r}
eur_index <- subset(rsimple, pop=="EUR" & fdr < 0.05)$prot
afr_index <- subset(rsimple, pop=="AFR" & fdr < 0.05)$prot
summary(lm(AFR ~ EUR, rw, subset=rw$prot %in% eur_index))
summary(lm(AFR ~ EUR, rw))
summary(lm(EAS ~ EUR, rw, subset=rw$prot %in% eur_index))
summary(lm(SAS ~ EUR, rw, subset=rw$prot %in% eur_index))
summary(lm(AFR ~ EUR, rw, subset=rw$prot %in% eur_index))

summary(lm(EUR ~ AFR, rw, subset=rw$prot %in% afr_index))
summary(lm(EUR ~ AFR, rw))

summary(lm(SAS ~ AFR, rw, subset=rw$prot %in% afr_index))
summary(lm(EAS ~ AFR, rw, subset=rw$prot %in% afr_index))
summary(lm(EAS ~ AFR, rw))

```

```{r}
rwm <- as.matrix(rw[,-1])
rownames(rwm) <- unlist(rw[,1])
heatmap(rwm)
```

Perform associations, adjusting for PCs, age etc

```{r}

phewas_covs <- function(prot4, dat) {
  lapply(names(prot4), function(anc) {
    x <- prot4[[anc]]
    x <- subset(x, id %in% dat$eid)
    ind <- match(x$id, dat$eid)
    d <- dat[ind, ]
    stopifnot(all(d$eid == x$id))
    lapply(4:ncol(x), function(i) {
      o <- summary(lm(x[[i]] ~ d$normscore + d$age_enrollment + d$sex + d$V3 + d$V4 + d$V5 + d$V6 + d$V7 + d$V8 + d$V9 + d$V10 + d$V11 + d$V12))
      tibble(pop = anc, prot = names(x)[i], bhat = o$coef[2,1], se = o$coef[2,2], pval = o$coef[2,4], n = o$df[1] + o$df[2])
    }) %>% bind_rows()
  }) %>% bind_rows()
}


rcovs <- phewas_covs(prot4, dat)
rcovs$fdr <- p.adjust(rcovs$pval, "fdr")
table(rcovs$fdr < 0.05, rcovs$pop)
```


```{r}
phewas_covs1 <- function(prot4, dat) {
  lapply(names(prot4), \(anc) {
    x <- prot4[[anc]]
    x <- subset(x, id %in% dat$eid)
    ind <- match(x$id, dat$eid)
    d <- dat[ind, ]
    stopifnot(all(d$eid == x$id))
    mclapply(4:ncol(x), \(i) {
      o <- summary(lm(x[[i]] ~ d$normscore + d$age_enrollment + d$sex + d$V3 + d$V4 + d$V5 + d$V6 + d$V7 + d$V8 + d$V9 + d$V10 + d$V11 + d$V12))
      tibble(pop=anc, prot=names(x)[i], bhat=o$coef[2,1], se=o$coef[2,2], pval=o$coef[2,4], n=o$df[1] + o$df[2])
    }) %>% bind_rows()
  }) %>% bind_rows()
}


rcovs1 <- phewas_covs1(prot4, dat)
rcovs1$fdr <- p.adjust(rcovs1$pval, "fdr")
table(rcovs1$fdr < 0.05, rcovs1$pop)
```


## Heterogeneity analysis

```{r}
fixed_effects_meta_analysis <- function(beta_vec, se_vec) {
    w <- 1 / se_vec^2
    beta <- sum(beta_vec * w, na.rm=T) / sum(w, na.rm=T)
    se <- sqrt(1 / sum(w, na.rm=T))
    pval <- pnorm(abs(beta / se), lower.tail = FALSE)
    Qj <- w * (beta-beta_vec)^2
    Q <- sum(Qj, na.rm=T)
    Qdf <- sum(!is.na(beta_vec))-1
    if(Qdf == 0) Q <- 0
    Qjpval <- pchisq(Qj, 1, lower.tail=FALSE)
    Qpval <- pchisq(Q, Qdf, lower.tail=FALSE)
    return(list(beta=beta, se=se, Q=Q, Qdf=Qdf, Qpval=Qpval, Qj=Qj, Qjpval=Qjpval))
}
```


##Do het analysis for all proteins

```{r}
rcovs_sig<- subset(rcovs1, fdr < 0.05)
head(rcovs_sig)
```

```{r}
table(rcovs_sig$pop)
```


```{r}
rcovs_sig2 <- subset(rcovs1, prot %in% rcovs_sig$prot)
table(rcovs_sig2$pop)
```

```{r}
subset(rcovs_sig2, prot == "acta2")
```


```{r}
fixed_effects_meta_analysis(subset(rcovs_sig2, prot == "acta2")$bhat, subset(rcovs_sig2, prot == "acta2")$se)
```

```{r}
temp <- subset(rcovs_sig2, prot == "acta2")
ggplot(temp, aes(x=bhat, y=pop)) +
  geom_point(aes(colour=pop)) +
  geom_errorbarh(aes(colour=pop, xmin=bhat - se * 1.96, xmax = bhat + se * 1.96), height=0) +
  geom_vline(xintercept=0, linetype="dotted")
```


```{r}
temp <- subset(rcovs_sig, prot == "acta2")
fixed_effects_meta_analysis(temp$bhat, temp$se)
ggplot(temp, aes(x=bhat, y=pop)) +
  geom_point(aes(colour=pop)) +
  geom_errorbarh(aes(colour=pop, xmin=bhat - se * 1.96, xmax = bhat + se * 1.96), height=0) +
  geom_vline(xintercept=0, linetype="dotted")

```

```{r}
res <- lapply(unique(rcovs_sig2$prot), \(e) {
  x <- subset(rcovs_sig2, prot == e)
  res <- fixed_effects_meta_analysis(x$bhat, x$se)
  Q <- tibble(exposure=e, beta=res$beta, se=res$se, Q=res$Q, Qdf=res$Qdf, Qpval=res$Qpval)
  Qj <- tibble(exposure=e, pop=x$pop, Qj=res$Qj, Qjpval=res$Qjpval)
  return(list(Q,Qj))
})

res[[1]]
```

```{r}
resQ <- lapply(res, \(x) x[[1]]) %>% bind_rows()
head(resQ)
```

```{r}
resQ$fdr <- p.adjust(resQ$Qpval, "fdr")
ggplot(resQ, aes(x=exposure, y=-log10(Qpval))) +
  geom_point() +
  geom_hline(yintercept=-log10(0.05/length(unique(rcovs_sig$prot))))
```

```{r}
resQ$fdr <- p.adjust(resQ$Qpval, "fdr")
ggplot(resQ, aes(x=exposure, y=-log10(fdr))) +
  geom_point() +
  geom_hline(yintercept=-log10(0.05))
```

```{r}
afr_sig <- subset(rcovs_sig, pop=="AFR")$prot

ggplot(resQ %>% filter(exposure %in% afr_sig), aes(x=exposure, y=-log10(Qpval))) +
  geom_point() +
  geom_hline(yintercept=-log10(0.05/length(unique(afr_sig))))
```


# Organise the results for per-study heterogeneity contributions (Qj)

```{r}
resQj <- lapply(res, \(x) x[[2]]) %>% bind_rows()

# Correct for multiple testing using FDR
resQj$fdr <- p.adjust(resQj$Qjpval, "fdr")
```


# Visualise the per-study heterogeneity (pval)
```{r}
ggplot(resQj, aes(x=exposure, y=-log10(Qjpval))) +
geom_point(aes(colour=pop)) +
geom_hline(yintercept=-log10(0.05))

```



# Visualise the per-study heterogeneity (FDR)

```{r}
ggplot(resQj, aes(x=exposure, y=-log10(fdr))) +
geom_point(aes(colour=pop)) +
geom_hline(yintercept=-log10(0.05))
```



For each of AFR, SAS and EAS, select their significant proteins, and perform heterogeneity (like in rcovs_sig2), to see if there's any evidence that these are ancestry specific




Choosing only significant proteins in the African ancestry
```{r}
selected_pop <- rcovs_sig %>%
  filter(pop %in% c("AFR")) %>%
  select(prot)

dim(selected_pop)

```

Obtaining protein information for all

```{r}
selected_pop_all <- subset(rcovs1, prot %in% selected_pop$prot)
table(selected_pop_all$pop)
```


Do heterogeneity analysis for all significant proteins in African ancestry for all ancestries

```{r}
res <- lapply(unique(selected_pop_all$prot), \(e) {
  x <- subset(selected_pop_all, prot == e)
  res <- fixed_effects_meta_analysis(x$bhat, x$se)
  Q <- tibble(exposure=e, beta=res$beta, se=res$se, Q=res$Q, Qdf=res$Qdf, Qpval=res$Qpval)
  Qj <- tibble(exposure=e, pop=x$pop, Qj=res$Qj, Qjpval=res$Qjpval)
  return(list(Q,Qj))
})

resQ <- lapply(res, \(x) x[[1]]) %>% bind_rows()


# Organise the results for per-study heterogeneity contributions (Qj)

resQj <- lapply(res, \(x) x[[2]]) %>% bind_rows()

# Correct for multiple testing using FDR
resQj$fdr <- p.adjust(resQj$Qjpval, "fdr")

res_m <- inner_join(selected_pop_all, resQj, by=c("prot"="exposure", "pop"))


names(resQ)[names(resQ)=="beta"] <- "b"
resQ$pop <- "Combined"
resQ$Qjpval <- 1


# Define a custom color palette
custom_colors <- c("SAS" = "#D32F2F", "EUR" = "#1976D2", "AFR" = "#388E3C", "EAS" = "#FBC02D")



ggplot(res_m, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(shape=Qjpval < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of African significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors)

```


Choosing only significant proteins in the EAS ancestry
```{r}
selected_pop <- rcovs_sig %>%
  filter(pop %in% c("EAS")) %>%
  select(prot)

dim(selected_pop)

```

Obtaining protein information for all

```{r}
selected_pop_all <- subset(rcovs1, prot %in% selected_pop$prot)
table(selected_pop_all$pop)
```


Do heterogeneity analysis for all significant proteins in EAS ancestry for all ancestries

```{r}
res <- lapply(unique(selected_pop_all$prot), \(e) {
  x <- subset(selected_pop_all, prot == e)
  res <- fixed_effects_meta_analysis(x$bhat, x$se)
  Q <- tibble(exposure=e, beta=res$beta, se=res$se, Q=res$Q, Qdf=res$Qdf, Qpval=res$Qpval)
  Qj <- tibble(exposure=e, pop=x$pop, Qj=res$Qj, Qjpval=res$Qjpval)
  return(list(Q,Qj))
})

resQ <- lapply(res, \(x) x[[1]]) %>% bind_rows()



# Organise the results for per-study heterogeneity contributions (Qj)

resQj <- lapply(res, \(x) x[[2]]) %>% bind_rows()

# Correct for multiple testing using FDR
resQj$fdr <- p.adjust(resQj$Qjpval, "fdr")

res_m <- inner_join(selected_pop_all, resQj, by=c("prot"="exposure", "pop"))


names(resQ)[names(resQ)=="beta"] <- "b"
resQ$pop <- "Combined"
resQ$Qjpval <- 1


# Define a custom color palette
custom_colors <- c("SAS" = "#D32F2F", "EUR" = "#1976D2", "AFR" = "#388E3C", "EAS" = "#FBC02D")



ggplot(res_m, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(size=Qjpval < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of EAS significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors) +
  scale_size_manual(
    values=c("FALSE"=2, "TRUE"=4),
    guide=guide_legend(title="Heterogeneity",
                       override.aes=list(shape=c(16,16))))

```



Choosing only significant proteins in the SAS ancestry
```{r}
selected_pop <- rcovs_sig %>%
  filter(pop %in% c("SAS")) %>%
  select(prot)

dim(selected_pop)

```

Obtaining protein information for all

```{r}
selected_pop_all <- subset(rcovs1, prot %in% selected_pop$prot)
table(selected_pop_all$pop)
```


Do heterogeneity analysis for all significant proteins in SAS ancestry for all ancestries

```{r, fig.width=10, fig.height=8}
res <- lapply(unique(selected_pop_all$prot), \(e) {
  x <- subset(selected_pop_all, prot == e)
  res <- fixed_effects_meta_analysis(x$bhat, x$se)
  Q <- tibble(exposure=e, beta=res$beta, se=res$se, Q=res$Q, Qdf=res$Qdf, Qpval=res$Qpval)
  Qj <- tibble(exposure=e, pop=x$pop, Qj=res$Qj, Qjpval=res$Qjpval)
  return(list(Q,Qj))
})

resQ <- lapply(res, \(x) x[[1]]) %>% bind_rows()



# Organise the results for per-study heterogeneity contributions (Qj)

resQj <- lapply(res, \(x) x[[2]]) %>% bind_rows()

# Correct for multiple testing using FDR
resQj$fdr <- p.adjust(resQj$Qjpval, "fdr")

res_m <- inner_join(selected_pop_all, resQj, by=c("prot"="exposure", "pop"))


names(resQ)[names(resQ)=="beta"] <- "b"
resQ$pop <- "Combined"
resQ$Qjpval <- 1


# Define a custom color palette
custom_colors <- c("SAS" = "#D32F2F", "EUR" = "#1976D2", "AFR" = "#388E3C", "EAS" = "#FBC02D")



ggplot(res_m, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(shape=Qjpval < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of SAS significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors)

```


Obtaining protein names that are significant in EAS, SAS and AFR
```{r}
selected_pop <- rcovs_sig %>%
  filter(pop %in% c("AFR", "SAS", "EAS")) %>%
  select(prot)

dim(selected_pop)

```
Obtaining protein information for all

```{r}
selected_pop_all <- subset(rcovs1, prot %in% selected_pop$prot)
table(selected_pop_all$pop)
```


Do heterogeneity analysis for all significant proteins in SAS ancestry for all ancestries

```{r, fig.width=12, fig.height=16}
res <- lapply(unique(selected_pop_all$prot), \(e) {
  x <- subset(selected_pop_all, prot == e)
  res <- fixed_effects_meta_analysis(x$bhat, x$se)
  Q <- tibble(exposure=e, beta=res$beta, se=res$se, Q=res$Q, Qdf=res$Qdf, Qpval=res$Qpval)
  Qj <- tibble(exposure=e, pop=x$pop, Qj=res$Qj, Qjpval=res$Qjpval)
  return(list(Q,Qj))
})

resQ <- lapply(res, \(x) x[[1]]) %>% bind_rows()



# Organise the results for per-study heterogeneity contributions (Qj)

resQj <- lapply(res, \(x) x[[2]]) %>% bind_rows()

# Correct for multiple testing using FDR
resQj$fdr <- p.adjust(resQj$Qjpval, "fdr")

res_m <- inner_join(selected_pop_all, resQj, by=c("prot"="exposure", "pop"))


names(resQ)[names(resQ)=="beta"] <- "b"
resQ$pop <- "Combined"
resQ$Qjpval <- 1


# Define a custom color palette
custom_colors <- c("SAS" = "#D32F2F", "EUR" = "#1976D2", "AFR" = "#388E3C", "EAS" = "#FBC02D")



ggplot(res_m, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(size=fdr.y < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of all significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors)

```

The forest plot is too long, the following analysis is for per ancestry significant proteins for all across all ancestries except EUR.


Choosing only significant proteins in the SAS ancestry
```{r}
selected_pop_SAS <- rcovs_sig %>%
  filter(pop %in% "SAS") %>%
  select(prot)

dim(selected_pop_SAS)

```

Obtaining protein information for all

```{r}
selected_pop_SAS_all <- subset(res_m, prot %in% selected_pop_SAS$prot)
table(selected_pop_SAS_all$pop)
```

```{r, fig.width=12, fig.height=14}
ggplot(selected_pop_SAS_all, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(size =fdr.y < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of SAS significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors) +
  scale_size_manual(
    values=c("FALSE"=2, "TRUE"=4),
    guide=guide_legend(title="Heterogeneity (FDR)",
                       override.aes=list(shape=c(16,16))))

```




Choosing only significant proteins in the EAS ancestry
```{r}
selected_pop_eas <- rcovs_sig %>%
  filter(pop %in% "EAS") %>%
  select(prot)

dim(selected_pop_eas)

```

Obtaining protein information for all

```{r}
selected_pop_eas_all <- subset(res_m, prot %in% selected_pop_eas$prot)
table(selected_pop_eas_all$pop)
```

```{r}
ggplot(selected_pop_eas_all, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(size =fdr.y < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of EAS significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors) +
  scale_size_manual(
    values=c("FALSE"=2, "TRUE"=4),
    guide=guide_legend(title="Heterogeneity (FDR)",
                       override.aes=list(shape=c(16,16))))

```


Choosing only significant proteins in the EAS ancestry
```{r}
selected_pop_afr <- rcovs_sig %>%
  filter(pop %in% "AFR") %>%
  select(prot)

dim(selected_pop_afr)

```

Obtaining protein information for all

```{r}
selected_pop_afr_all <- subset(res_m, prot %in% selected_pop_afr$prot)
table(selected_pop_afr_all$pop)
```


```{r}
ggplot(selected_pop_afr_all, aes(y=pop, x=bhat, color=pop)) +
  geom_point(aes(size =fdr.y < 0.05)) +
  geom_errorbarh(aes(xmin=bhat-se*1.96, xmax=bhat+se*1.96), height=0) +
  geom_vline(xintercept=0, linetype="dashed") +
  labs(title = "Forest plot of AFR significant proteins across all ancestries",
       x = "beta",
       y = "Proteins") +
   facet_grid(prot ~ .)+
   theme_minimal() +  # Apply a minimal theme
  theme(
    plot.title = element_text(hjust = 0.5), # Center the plot title
    axis.text.y=element_blank(), # remove y axis text
    strip.text.y=element_text(angle=0)
  ) +
  scale_colour_manual(values=custom_colors) +
  scale_size_manual(
    values=c("FALSE"=2, "TRUE"=4),
    guide=guide_legend(title="Heterogeneity (FDR)",
                       override.aes=list(shape=c(16,16))))

```




## Expected vs observed replication

Take significant results from EUR
Even though the power is lower in other populations, is the degree of association high enough to suggest that overall, the effects are the same

```{r}
evso <- function(b_disc, b_rep, se_disc, se_rep, alpha)
{
  p_sign <- pnorm(-abs(b_disc) / se_disc) * pnorm(-abs(b_disc) / se_rep) + ((1 - pnorm(-abs(b_disc) / se_disc)) * (1 - pnorm(-abs(b_disc) / se_rep)))
  p_sig <- pnorm(-abs(b_disc) / se_rep + qnorm(alpha / 2)) + (1 - pnorm(-abs(b_disc) / se_rep - qnorm(alpha / 2)))
  p_rep <- pnorm(abs(b_rep)/se_rep, lower.tail=FALSE)
  res <- tibble::tibble(
    nsnp=length(b_disc),
    metric=c("Sign", "Sign", "P-value", "P-value"),
    datum=c("Expected", "Observed", "Expected", "Observed"),
    value=c(sum(p_sign, na.rm=TRUE), sum(sign(b_disc) == sign(b_rep)), sum(p_sig, na.rm=TRUE), sum(p_rep < alpha, na.rm=TRUE))
  )
  return(list(res=res, variants=dplyr::tibble(sig=p_sig, sign=p_sign)))
}
```

EUR vs AFR

```{r}
eur_prot <- subset(rcovs_sig, pop == "EUR")$prot
temp1 <- rcovs1[rcovs1$pop == "EUR" & rcovs1$prot %in% eur_prot, ]
temp2 <- rcovs1[rcovs1$pop == "AFR" & rcovs1$prot %in% eur_prot, ]
evso(b_disc = temp1$bhat, b_rep = temp2$bhat, se_disc = temp1$se, se_rep = temp2$se, 0.05)
```

AFR vs EUR

```{r}
afr_prot <- subset(rcovs_sig, pop == "AFR")$prot
temp1 <- rcovs1[rcovs1$pop == "AFR" & rcovs1$prot %in% afr_prot, ]
temp2 <- rcovs1[rcovs1$pop == "EUR" & rcovs1$prot %in% afr_prot, ]
evso(b_disc = temp1$bhat, b_rep = temp2$bhat, se_disc = temp1$se, se_rep = temp2$se, 0.05)

```


EUR vs SAS

```{r}
eur_prot <- subset(rcovs_sig, pop == "EUR")$prot
temp1 <- rcovs1[rcovs1$pop == "EUR" & rcovs1$prot %in% eur_prot, ]
temp2 <- rcovs1[rcovs1$pop == "SAS" & rcovs1$prot %in% eur_prot, ]
evso(b_disc = temp1$bhat, b_rep = temp2$bhat, se_disc = temp1$se, se_rep = temp2$se, 0.05)
```
SAS vs EUR

```{r}
sas_prot <- subset(rcovs_sig, pop == "SAS")$prot
temp1 <- rcovs1[rcovs1$pop == "SAS" & rcovs1$prot %in% sas_prot, ]
temp2 <- rcovs1[rcovs1$pop == "EUR" & rcovs1$prot %in% sas_prot, ]
evso(b_disc = temp1$bhat, b_rep = temp2$bhat, se_disc = temp1$se, se_rep = temp2$se, 0.05)

```



EUR vs EAS

```{r}
eur_prot <- subset(rcovs_sig, pop == "EUR")$prot
temp1 <- rcovs1[rcovs1$pop == "EUR" & rcovs1$prot %in% eur_prot, ]
temp2 <- rcovs1[rcovs1$pop == "EAS" & rcovs1$prot %in% eur_prot, ]
evso(b_disc = temp1$bhat, b_rep = temp2$bhat, se_disc = temp1$se, se_rep = temp2$se, 0.05)
```

EAS vs EUR
```{r}
eas_prot <- subset(rcovs_sig, pop == "EAS")$prot
temp1 <- rcovs1[rcovs1$pop == "EAS" & rcovs1$prot %in% eas_prot, ]
temp2 <- rcovs1[rcovs1$pop == "EUR" & rcovs1$prot %in% eas_prot, ]
evso(b_disc = temp1$bhat, b_rep = temp2$bhat, se_disc = temp1$se, se_rep = temp2$se, 0.05)

```

There is a higher observed replication in AFR, SAS, and EAS populations compared to what was expected based on EUR data.
meaning that the protein associations are also relevant in these populations.
