---
title: "Untitled"
author: "Linda"
date: "2024-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(here)
library(dplyr)
library(glmnet)
library(ggplot2)
library(caret)
library(tidyr)
```

Preparing the data
```{r}
load("C:/Users/HP OMEN GAMING/Documents/mdd-proteome-ancestry/data/ukb_prot_mdd.rdata")
```
create a column the categorizes cases into prospective cases and retro and controls
```{r}
dat <- dat %>%
  mutate(
    prospective.depression = case_when(
      depression == 0 & is.na(age_first_episode) ~ "control",
      !is.na(age_first_episode) & age_first_episode <= age_enrollment ~ "retro",
      !is.na(age_first_episode) & age_first_episode > age_enrollment ~ "pros",
      TRUE ~ NA_character_
    )
  )
table(dat$prospective.depression)

```
control    MDD- pros   MDD-retro 
  42359    1190        7908 
  
  
merge new data with prot data and replace NA value with mean 
```{r}
temp <- inner_join(dat %>% select(eid, prospective.depression, depression),
    prot4 %>% bind_rows(), by=c("eid"="id"))

#temp <- temp %>% select(-eid, -pop)

# Set values in temp2 to mean if NA
temp <- temp %>%
  mutate(across(where(is.numeric), ~replace_na(., mean(., na.rm = TRUE))))
temp$prospective.depression<- as.factor(temp$prospective.depression)
temp$depression<- as.factor(temp$depression)
summary(temp[,1:10])
```
Visulizations
Distributions of PRS in the MDD caseS and controls
```{r}
tibble(dep=temp$depression, PRS=temp$normscore) %>%
  ggplot(aes(y=PRS, x=dep)) +
  geom_boxplot()
```

Distributions of PRS in the MDD prospective and prevalent caseS and controls
```{r}
tibble(dep=temp$prospective.depression, PRS=temp$normscore) %>%
  ggplot(aes(y=PRS, x=dep)) +
  geom_boxplot()
```

Distributions of PRS in the MDD caseS and controls per ancestry
```{r}
ggplot(temp, aes(x = pop, y = normscore, fill = depression)) +
  geom_boxplot() +
  labs(title = "Box Plot of MDD PRS by Population",
       x = "Population Group",
       y = "MDD PRS",
       fill = "Depression Status (0=Control, 1=Case)") +
  theme_minimal()

```

Distributions of PRS in the prospective MDD caseS and controls per ancestry
```{r}
ggplot(temp, aes(x = pop, y = normscore, fill = prospective.depression)) +
  geom_boxplot() +
  labs(title = "Box Plot of MDD PRS by Population",
       x = "Population Group",
       y = "MDD PRS",
       fill = "Prospective depression Status") +
  theme_minimal()

```

```{r}
temp %>%
  group_by(pop, depression) %>%
  summarise(count = n(), .groups = 'drop')
```

```{r}
table(temp$depression)
```




Data split
train = all ancestries  (no prospective cases + 87% controls)
test = all ancestries (all prospective cases + 13% controls)
```{r}
set.seed(123)  # For reproducibility

# Split control data into 67% for training and 33% for testing
control_indices <- which(temp$prospective.depression == "control")
train_control_indices <- sample(control_indices, size = 0.87 * length(control_indices))
test_control_indices <- setdiff(control_indices, train_control_indices)

# Get indices for all "retro" cases (all go to training set)
retro_indices <- which(temp$prospective.depression == "retro")

# Get indices for all "pros" cases (all go to testing set)
pros_indices <- which(temp$prospective.depression == "pros")

# Create training data (all retro + 87% controls)
train_data_initial <- temp[c(train_control_indices, retro_indices), ]

# Create testing data (all pros + 13% controls)
test_data_initial <- temp[c(test_control_indices, pros_indices), ]

# Check the distribution in the training and testing sets
table(train_data_initial$prospective.depression)
table(test_data_initial$prospective.depression)

```

remove prospective.depression colummn to remain with only the required columns
```{r}
test_data <- test_data_initial %>% select(-prospective.depression, -eid)
train_data <- train_data_initial %>% select(-prospective.depression, -eid,)

```

```{r}
test_data_pro<-test_data %>% select(4:ncol(test_data))
train_data_pro<-train_data %>% select(4:ncol(train_data))
```

Fit the model with cross validation
```{r}
set.seed(123)  # For reproducibility
cvres2 <- cv.glmnet(
    x = as.matrix(train_data_pro),
    y = train_data$normscore,
    alpha = 0.5,
    nfolds = 5
)

```


```{r}
# Look at the performance of the model for different values of lambda
plot(cvres2)
```

Generate the predictor in the All ancestry test set
```{r}

# This uses the set of coefficients that has the minimum lambda value from the cross validation
pred2 <- predict(cvres2, newx = as.matrix(test_data_pro), s = "lambda.min")
```

```{r}
hist(pred2, breaks=100)
```


```{r}
temp_data <- tibble(y=test_data$depression, x=pred2)
all_roc_result<-pROC::roc(y ~ x, temp_data)
print(all_roc_result)

# Extract the AUC value
all_auc_value3 <- all_roc_result$auc
print(all_auc_value3) 
```


Testing in the African data set
```{r}
# Filter for rows where the population is 'AFR'
afr_test_data <- test_data %>% filter(pop == "AFR")
#afr protein data
afr_test_pro<-afr_test_data%>% select(4:ncol(afr_test_data))
#predictions
afr_pred2 <- predict(cvres2, newx = as.matrix(afr_test_pro), s = "lambda.min")
hist(afr_pred2, breaks=100)
# Model performance metrics
afr_temp_data <- tibble(y=afr_test_data$depression, x=afr_pred2)
# Create the ROC object
afr_roc_result <- pROC::roc(y ~ x, data = afr_temp_data)
print(afr_roc_result)

# Extract the AUC value
afr_auc_value3 <- afr_roc_result$auc
print(afr_auc_value3) 
```

Testing in the EUR data set
```{r}
# Filter for rows where the population is 'EUR'
eur_test_data <- test_data %>% filter(pop == "EUR")
#eur protein data
eur_test_pro<-eur_test_data%>% select(4:ncol(eur_test_data))
#predictions
eur_pred2 <- predict(cvres2, newx = as.matrix(eur_test_pro), s = "lambda.min")
hist(eur_pred2, breaks=100)
# Model performance metrics
eur_temp_data <- tibble(y=eur_test_data$depression, x=eur_pred2)
# Create the ROC object
eur_roc_result <- pROC::roc(y ~ x, data = eur_temp_data)
print(eur_roc_result)

# Extract the AUC value
eur_auc_value3 <- eur_roc_result$auc
print(eur_auc_value3) 
```
Testing in the SAS data set
```{r}
# Filter for rows where the population is 'SAS'
sas_test_data <- test_data %>% filter(pop == "SAS")
#sas protein data
sas_test_pro<-sas_test_data%>% select(4:ncol(sas_test_data))
#predictions
sas_pred2 <- predict(cvres2, newx = as.matrix(sas_test_pro), s = "lambda.min")
hist(sas_pred2, breaks=100)
# Model performance metrics
sas_temp_data <- tibble(y=sas_test_data$depression, x=sas_pred2)
# Create the ROC object
sas_roc_result <- pROC::roc(y ~ x, data = sas_temp_data)
print(sas_roc_result)

# Extract the AUC value
sas_auc_value3 <- sas_roc_result$auc
print(sas_auc_value3) 
```

Testing in the EAS data set
```{r}

# Filter for rows where the population is 'EAS'
eas_test_data <- test_data %>% filter(pop == "EAS")
#eas protein data
eas_test_pro<-eas_test_data%>% select(4:ncol(eas_test_data))
#predictions
eas_pred2 <- predict(cvres2, newx = as.matrix(eas_test_pro), s = "lambda.min")
hist(eas_pred2, breaks=100)
# Model performance metrics
eas_temp_data <- tibble(y=eas_test_data$depression, x=eas_pred2)
# Create the ROC object
eas_roc_result <- pROC::roc(y ~ x, data = eas_temp_data)
print(eas_roc_result)

# Extract the AUC value
eas_auc_value3 <- eas_roc_result$auc
print(eas_auc_value3) 
```


## (2)protein -> MDD prospective cases (all ancestries)

Remain with only prospective cases and controls
```{r}
temp2 <- temp %>% filter(prospective.depression !="retro")
temp2$prospective.depression <- droplevels(temp2$prospective.depression)
table(temp2$prospective.depression)
```

MDD status per ancestry in the data
```{r}
temp2 %>%
  group_by(pop, depression) %>%
  summarise(count = n(), .groups = 'drop')
```


Data Split
```{r}
set.seed(123)  # For reproducibility
ind<-createDataPartition(temp2$depression, p = 0.67, list = FALSE)
```

Data preparation
```{r}
temp_2 <- temp2 %>% select(-prospective.depression, -eid)
temp1 <- temp_2[,1:2]

temp2<-temp_2 %>% select(-pop, -depression, -normscore )
```


# Fit the model with cross validation
# The cross validation will help to choose the best lambda value (i.e. how much shrinkage does best for prediction)
```{r}
set.seed(123)
cvres <- cv.glmnet(
    x = as.matrix(temp2[ind,]),
    y = temp1$depression[ind],
    alpha = 0.5,
    family = "binomial",
    nfolds = 5
)




```

```{r}
plot(cvres)
```

```{r}
cvres$lambda.min
cvres$lambda.1se

```

Computing the final model
```{r}
# Final model with lambda.min
net.model <- glmnet(
  x = as.matrix(temp2[ind,]),
  y = temp1$depression[ind],
  alpha = 0.5,
  family = "binomial",
  lambda = cvres$lambda.min
)


```


Making predictions on all ancestry test data

Predicting probabilities that each observation belongs to the positive class (class 1)
```{r}
all_pred<-net.model %>% predict(newx=as.matrix(temp2[-ind,]))
hist(all_pred, breaks=100)
```

```{r}
library(pROC)
all_data <- tibble(y=temp1$depression[-ind], x=all_pred)
all_roc_result2<- pROC::roc(y ~ x, all_data)
print(all_roc_result2)

# Extract the AUC value
all_auc_value2 <- all_roc_result2$auc
print(all_auc_value2) 
```

Testing in AFR population
```{r}
test_2 <- temp_2[-ind, ]
afr_test_data2<-test_2 %>% filter(pop == "AFR")
afr_temp1.2 <- afr_test_data2[,1:2]
afr_temp2.2<-afr_test_data2 %>% select(-pop, -depression, -normscore)

#Making predictions on afr test data
afr_pred<-net.model %>% predict(newx=as.matrix(afr_temp2.2))
hist(afr_pred, breaks=100)

afr_data <- tibble(y=afr_temp1.2$depression, x=afr_pred)
afr_roc_result2<-pROC::roc(y ~ x, afr_data)
print(afr_roc_result2)

# Extract the AUC value
afr_auc_value2 <- afr_roc_result2$auc
print(afr_auc_value2) 


```

Testing in EUR population
```{r}
#test_2 <- temp_2[-ind, ]
eur_test_data2<-test_2 %>% filter(pop == "EUR")
eur_temp1.2 <- eur_test_data2[,1:2]
eur_temp2.2<-eur_test_data2 %>% select(-pop, -depression, -normscore)

#Making predictions on eur test data
eur_pred<-net.model %>% predict(newx=as.matrix(eur_temp2.2))
hist(eur_pred, breaks=100)

eur_data <- tibble(y=eur_temp1.2$depression, x=eur_pred)
eur_roc_result2<-pROC::roc(y ~ x, eur_data)
print(eur_roc_result2)

# Extract the AUC value
eur_auc_value2 <- eur_roc_result2$auc
print(afr_auc_value2) 


```


Testing in SAS population
```{r}
#test_2 <- temp_2[-ind, ]
sas_test_data2<-test_2 %>% filter(pop == "SAS")
sas_temp1.2 <- sas_test_data2[,1:2]
sas_temp2.2<-sas_test_data2 %>% select(-pop, -depression, -normscore)

#Making predictions on sas test data
sas_pred<-net.model %>% predict(newx=as.matrix(sas_temp2.2))
hist(sas_pred, breaks=100)

sas_data <- tibble(y=sas_temp1.2$depression, x=sas_pred)
sas_roc_result2<-pROC::roc(y ~ x, sas_data)
print(sas_roc_result2)

# Extract the AUC value
sas_auc_value2 <- sas_roc_result2$auc
print(sas_auc_value2) 


```



Testing in EAS population
```{r}
#test_2 <- temp_2[-ind, ]
eas_test_data2<-test_2 %>% filter(pop == "EAS")
eas_temp1.2 <- eas_test_data2[,1:2]
eas_temp2.2<-eas_test_data2 %>% select(-pop, -depression, -normscore)

#Making predictions on eas test data
eas_pred<-net.model %>% predict(newx=as.matrix(eas_temp2.2))
hist(eas_pred, breaks=100)

eas_data <- tibble(y=eas_temp1.2$depression, x=eas_pred)
eas_roc_result2<-pROC::roc(y ~ x, eas_data)
print(eas_roc_result2)

# Extract the AUC value
eas_auc_value2 <- eas_roc_result2$auc
print(eas_auc_value2) 


```
Compute AUC and CI for both reverse MR and observational prediction for each ancestry
```{r}
# Reverse MR prediction
all_auc_ci_reverse <- pROC::ci.auc(all_roc_result)
sas_auc_ci_reverse <- pROC::ci.auc(sas_roc_result)
eur_auc_ci_reverse <- pROC::ci.auc(eur_roc_result)
afr_auc_ci_reverse <- pROC::ci.auc(afr_roc_result)
eas_auc_ci_reverse <- pROC::ci.auc(eas_roc_result)

# Observational prediction
all_auc_ci_obs <- pROC::ci.auc(all_roc_result2)
sas_auc_ci_obs <- pROC::ci.auc(sas_roc_result2)
eur_auc_ci_obs <- pROC::ci.auc(eur_roc_result2)
afr_auc_ci_obs <- pROC::ci.auc(afr_roc_result2)
#eas_auc_value2 <- 0  # For observational, no cases for EAS
eas_auc_ci_obs <- pROC::ci.auc(eas_roc_result2)

```

Creating a Data Frame with AUC Values and CIs

```{r}
auc_data_ci <- data.frame(
  Ancestry = rep(c("All Ancestry", "SAS", "EUR", "AFR", "EAS"), each = 2),
  AUC_Type = rep(c("Reverse MR Prediction", "Observational Prediction"), times = 5),
  AUC = c(all_auc_value3, all_auc_value2, sas_auc_value3, sas_auc_value2, 
          eur_auc_value3, eur_auc_value2, afr_auc_value3, afr_auc_value2, 
          eas_auc_value3, eas_auc_value2),
  Lower_CI = c(all_auc_ci_reverse[1], all_auc_ci_obs[1], sas_auc_ci_reverse[1], sas_auc_ci_obs[1], 
               eur_auc_ci_reverse[1], eur_auc_ci_obs[1], afr_auc_ci_reverse[1], afr_auc_ci_obs[1], 
               eas_auc_ci_reverse[1], eas_auc_ci_obs[1]),
  Upper_CI = c(all_auc_ci_reverse[3], all_auc_ci_obs[3], sas_auc_ci_reverse[3], sas_auc_ci_obs[3], 
               eur_auc_ci_reverse[3], eur_auc_ci_obs[3], afr_auc_ci_reverse[3], afr_auc_ci_obs[3], 
               eas_auc_ci_reverse[3], eas_auc_ci_obs[3])
)

# View the AUC data with CIs
print(auc_data_ci)
write.csv(auc_data_ci, "C:/Users/HP OMEN GAMING/Documents/mdd-proteome-ancestry/data/AUC Values.csv", row.names = FALSE)

```

Create a Forest Plot for Both AUC Types for each ancestry
```{r}

p <- ggplot(auc_data_ci, aes(y = Ancestry, x = AUC, color = AUC_Type, group = AUC_Type)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(xmin = Lower_CI, xmax = Upper_CI), width = 0.2, position = position_dodge(width = 0.5)) +
  labs(title = "Forest Plot of AUC Values by Ancestry and Prediction Type",
       y = "Ancestry",
       x = "AUC (95% CI)",
       color = "AUC Type") +
  theme_bw() +
  theme(axis.text.y = element_text(angle = 45, hjust = 1))

# Save the plot as a PDF
ggsave(filename = "C:/Users/HP OMEN GAMING/Documents/mdd-proteome-ancestry/results/AUC-Values_ForestPlot.pdf", plot = p, width = 8, height = 6)

ggsave(filename = "C:/Users/HP OMEN GAMING/Documents/mdd-proteome-ancestry/results/AUC-Values_ForestPlot.png", plot = p, width = 8, height = 6)

```

```{r}

ggplot(auc_data_ci, aes(y = Ancestry, x = AUC, color = AUC_Type, group = AUC_Type)) +
  geom_point(size = 3, position = position_dodge(width = 0.5)) +
  geom_errorbar(aes(xmin = Lower_CI, xmax = Upper_CI), width = 0.2, position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black") +  # Add vertical dashed line at 0.5
  labs(title = "Forest Plot of AUC Values by Ancestry and AUC Type",
       y = "Ancestry",
       x = "AUC (95% CI)",
       color = "AUC Type") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 45, hjust = 1))  # Adjust text angle if needed

```

















