---
title: "Reverse_MR"
author: "Sandra"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
library(dplyr)
library(data.table)
library(ggplot2)
library(parallel)
library(glue)
library(tidyr)
library(tidyverse)
```

```{r}
ukb <-  load("ukb_prot_mdd.rdata")
Prelim <-  load("prelim-results.rdata")

```




```{r}
head(rsimple)
```

## Perform associations

```{r}
fast_assoc <- function(y, x) {
	index <- is.finite(y) & is.finite(x)
	n <- sum(index)
	y <- y[index]
	x <- x[index]
	#vx <- var(x)
	vy <- var(y)
	bhat <- stats::cov(y, x) / vy
	ahat <- mean(x) - bhat * mean(y)
	# fitted <- ahat + x * bhat
	# residuals <- y - fitted
	# SSR <- sum((residuals - mean(residuals))^2)
	# SSF <- sum((fitted - mean(fitted))^2)

	rsq <- (bhat * vy)^2 / (vy * var(x))
	fval <- rsq * (n-2) / (1-rsq)
	tval <- sqrt(fval)
	se <- abs(bhat / tval)

	# Fval <- (SSF) / (SSR/(n-2))
	# pval <- pf(Fval, 1, n-2, lowe=F)
	p <- stats::pf(fval, 1, n-2, lower.tail=FALSE)
	return(list(
		ahat=ahat, bhat=bhat, se=se, fval=fval, pval=p, n=n
	))
}

phewas <- function(y, g, logistic=FALSE) {
	out <- matrix(0, ncol(g), 6)
	if(logistic)
	{
		stopifnot(all(y %in% c(0,1)))
		for(i in 1:ncol(g))
		{
			o <- logistic_assoc(y, g[,i])
			out[i, ] <- unlist(o)
		}
	} else {
		for(i in 1:ncol(g))
		{
			o <- fast_assoc(y, g[,i])
			out[i, ] <- unlist(o)
		}
	}

	out <- dplyr::as_tibble(out, .name_repair="minimal")
	names(out) <- names(o)
	out$snp <- 1:ncol(g)
	return(out)
}
```

```{r}
pops <- c("AFR", "EUR", "SAS", "EAS")
```

Perform simple analysis - no covariates etc

```{r}
rsimple <- lapply(pops, \(p) {
  x <- prot4[[p]]
  o <- phewas(y=unlist(x$normscore), g=as.matrix(as.matrix(x[,4:ncol(x)])))
  o$pop=p
  o$prot <- colnames(x[4:ncol(x)])
  o$fdr <- p.adjust(o$pval, "fdr")
  return(o)
}) %>% bind_rows()

rsimple %>% group_by(pop) %>%
  summarise(n=sum(fdr < 0.05))
```

Try comparing across ancestries

```{r}
rw <- spread(rsimple %>% select(pop, bhat, prot), key=pop, value=bhat)
head(rw)
```
# Subsetting the r simple dataset to obtain a dataframe with European and Africans protein with fdr < 0.05

```{r}
eur_index <- subset(rsimple, pop=="EUR" & fdr < 0.05)$prot
afr_index <- subset(rsimple, pop=="AFR" & fdr < 0.05)$prot
```

# Performing Linear regression 
```{r}
summary(lm(AFR ~ EUR, rw, subset=rw$prot %in% eur_index))
summary(lm(AFR ~ EUR, rw))
summary(lm(EAS ~ EUR, rw, subset=rw$prot %in% eur_index))
summary(lm(SAS ~ EUR, rw, subset=rw$prot %in% eur_index))
summary(lm(AFR ~ EUR, rw, subset=rw$prot %in% eur_index))

summary(lm(EUR ~ AFR, rw, subset=rw$prot %in% afr_index))
summary(lm(EUR ~ AFR, rw))

summary(lm(SAS ~ AFR, rw, subset=rw$prot %in% afr_index))
summary(lm(EAS ~ AFR, rw, subset=rw$prot %in% afr_index))
summary(lm(EAS ~ AFR, rw))
```
# Heatmap showing ancestry comparison
```{r}
rwm <- as.matrix(rw[,-1])
rownames(rwm) <- unlist(rw[,1])
heatmap(rwm)
```
Perform associations, adjusting for PCs, age etc

```{r}
phewas_covs <- function(prot4, dat) {
  lapply(names(prot4), \(anc) {
    x <- prot4[[anc]]
    x <- subset(x, id %in% dat$eid)
    ind <- match(x$id, dat$eid)
    d <- dat[ind, ]
    stopifnot(all(d$eid == x$id))
    mclapply(4:ncol(x), \(i) {
      o <- summary(lm(x[[i]] ~ d$normscore + d$age_enrollment + d$sex + d$V3 + d$V4 + d$V5 + d$V6 + d$V7 + d$V8 + d$V9 + d$V10 + d$V11 + d$V12))
      tibble(pop=anc, prot=names(x)[i], bhat=o$coef[2,1], se=o$coef[2,2], pval=o$coef[2,4], n=o$df[1] + o$df[2])
    }) %>% bind_rows()
  }) %>% bind_rows()
}

```

```{r}
rcovs <- phewas_covs(prot4, dat)
rcovs$fdr <- p.adjust(rcovs$pval, "fdr")
table(rcovs$fdr < 0.05, rcovs$pop)


```

```{r}
rwc <- spread(rcovs %>% select(pop, bhat, prot), key=pop, value=bhat)
rwc
```

Correlation of mddprs-prot assocs across ancestries

```{r}
cor(rwc[,-1], use="pair")
```

regression of mddprs-prot assocs across ancestries, using all proteins, or just those that were FDR < 0.05 in the discovery ancestry

```{r}
comps <- expand.grid(pop1=pops, pop2=pops, fdr=c(0.05, 1), b=NA, se=NA, pval=NA, n=NA)
comps <- lapply(1:nrow(comps), \(i) {
  ind <- subset(rcovs, pop==comps$pop1[i]) %>%
    filter(fdr < comps$fdr[i]) %>%
    {.$prot}
  if(length(ind) > 20) {
    o <- summary(lm(rwc[[comps$pop1[[i]]]][rwc$prot %in% ind] ~ rwc[[comps$pop2[[i]]]][rwc$prot %in% ind]))
    comps$b[i] <- o$coef[2,1]
    comps$se[i] <- o$coef[2,2]
    comps$pval[i] <- o$coef[2,4]
    comps$n[i] <- length(ind)
  }
  return(comps[i,])
}) %>% bind_rows()
comps %>%
  filter(pop1 != pop2) %>%
  ggplot(., aes(y=b, x=pop2)) +
  geom_point(aes(size=-log10(pval))) +
  facet_grid(pop1 ~ as.factor(fdr))

```
## Excluding cases


# Creating a new column with cases and controls for depression
```{r}
# Create a new column named 'Case_Control'
dat <- dat %>%
  mutate(Case_Control = ifelse(depression == 1, "Case", "Control"))

table(dat$Case_Control)

```

# Group by POP (ancestry) and calculate the count of cases and controls
```{r}
pop_counts <- dat %>%
  group_by(pop, Case_Control) %>%
  summarize(Count = n())

print(pop_counts)
 
#Convert Depression to a factor with meaningful labels
pop_counts$Case_Control<- factor(pop_counts$Case_Control, labels = c("Cases", "Control"))

# Create a bar plot
ggplot(pop_counts, aes(x = pop, y = Count, fill = Case_Control)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(title = "Cases and Controls per Ancestry",
       x = "Ancestry (Population)",
       y = "Frequency") +
  scale_fill_manual(values = c("Cases" = "blue", "Control" = "red")) +
  theme_minimal()

```
```{r}
covariate_columns <- c("depression", "normscore", "age_enrollment" ,
                       "V3", "V4", "V5")

# Subset the data to include only the selected covariates
covariate_data <- dat[, ..covariate_columns, drop = FALSE]

# Filter numeric columns only
numeric_columns <- names(Filter(is.numeric, covariate_data))
covariate_data <- covariate_data[, ..numeric_columns, drop = FALSE]

# Check if there are numeric columns
if (ncol(covariate_data) == 0) {
  stop("No numeric covariates found.")
}

# Calculate the correlation matrix
cor_matrix <- cor(covariate_data)

# Print the correlation matrix
print(cor_matrix)

# Visualization using a heatmap
library(ggplot2)
library(reshape2)

# Melt the correlation matrix for plotting
melted_cor_matrix <- melt(cor_matrix)

# Plot the heatmap
ggplot(melted_cor_matrix, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(
    title = "Correlation Heatmap for Covariates",
    x = "Covariate",
    y = "Covariate"
  )



```


# Obtaining the proteins that are ancestry specific and also performing a linear regression to understand the relationship between the proteins and depression (Cases and controls ) while adjusting for covariates like sex, normscore and age_enrolnmenr. ? THis can be adjusted further to adjust for the PCs 

```{r}

phewas_covs_1 <- function(prot4, dat) {
  lapply(names(prot4), \(anc) {
    x <- prot4[[anc]]
    x <- subset(x, id %in% dat$eid)
    ind <- match(x$id, dat$eid)
    d <- dat[ind, ]
    stopifnot(all(d$eid == x$id))
    mclapply(4:ncol(x), \(i) {
      #o <- summary(lm(x[[i]] ~ d$depression * d$sex + d$age_enrollment + d$V3 + d$V4 + d$V5 + d$V6 + d$V7 + d$V8 + d$V9 + d$V10 + d$V11 + d$V12))
      o <- summary(lm(x[[i]] ~ d$depression ))
      tibble(pop=anc, prot=names(x)[i], bhat=o$coef[2,1], se=o$coef[2,2], pval=o$coef[2,4], n=o$df[1] + o$df[2])
    }) %>% bind_rows()
  }) %>% bind_rows()
}


```

# Obtaining a count of proteins per ancestry with fdr less than 0.05 as either true or false 
```{r}
rcovs_1 <- phewas_covs_1(prot4, dat)
rcovs_1$fdr <- p.adjust(rcovs_1$pval, "fdr")
table(rcovs_1$fdr < 0.05, rcovs_1$pop)


```
#Extracting the proteins that are specific to Africans with fdr less than 0.05 

```{r}
# Extract proteins for a specific ancestry with FDR < 0.05
significant_proteins_ancestry <- rcovs_1$prot[rcovs_1$fdr < 0.05 & rcovs_1$pop == "AFR"]

# Display the significant proteins for the ancestry
print(significant_proteins_ancestry)

```

# Extract all proteins that are significantly associated with FDR < 0.05 per ancestry 
```{r}
# Extract significant proteins for all ancestries with FDR < 0.05
significant_proteins_all_ancestries <- rcovs_1 %>%
  filter(fdr < 0.05) %>%
  select(pop, prot)

# Display the table of significant proteins for all ancestries
print(significant_proteins_all_ancestries)

```


# Obtaining a list of proteins that are shared by more than 1 ancestry 
```{r}
# Set the minimum number of ancestries for proteins to be considered shared
min_ancestry_count <- 2  # Change this number based on your criteria

# Count the total number of ancestries
total_ancestry_count <- n_distinct(significant_proteins_all_ancestries$pop)

# Group by protein and count the number of unique ancestries for each protein
proteins_shared_across_ancestries <- significant_proteins_all_ancestries %>%
  group_by(prot) %>%
  summarise(
    unique_ancestry_count = n_distinct(pop),
    unique_ancestries = paste(unique(pop), collapse = ", ")
  )

# Filter proteins that are significant in more than one ancestry
shared_proteins <- proteins_shared_across_ancestries %>%
  filter(unique_ancestry_count >= min_ancestry_count)

# Display the table of shared proteins and their associated ancestries
print(shared_proteins)

```

```{r}
library(ggplot2)

# Plot shared proteins and their associated ancestries
ggplot(shared_proteins, aes(x = unique_ancestry_count, y = prot, fill = unique_ancestries)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Shared Proteins Across Ancestries",
    x = "Count",
    y = "Proteins"
  ) +
  theme_minimal()


```


# Obtaining proteins for prospective MDD cases 

```{r}
dat$prospective <- dat$age_first_episode > dat$age_enrollment
table(dat$prospective, dat$pop)
```

```{r}

prospective_dat <- dat[complete.cases(dat$prospective), ]
head(prospective_dat)
```


# The true is converted to 1 got MDD during the study, False had MDD during enrollment
# There are no proteins associated with the prospective column

```{r}
prospective_dat$prospective_num <- as.numeric(prospective_dat$prospective)
```

```{r}
phewas_covs_2 <- function(prot4, prospective_dat) {
  lapply(names(prot4), \(anc) {
    x <- prot4[[anc]]
    x <- subset(x, id %in% prospective_dat$eid)
    ind <- match(x$id, prospective_dat$eid)
    d <- prospective_dat[ind, ]
    stopifnot(all(d$eid == x$id))
    mclapply(4:ncol(x), \(i) {
      o <- summary(lm(x[[i]] ~ d$depression * d$sex + d$age_enrollment + d$V3 + d$V4 + d$V5 + d$V6 + d$V7 + d$V8 + d$V9 + d$V10 + d$V11 + d$V12))
      tibble(pop=anc, prot=names(x)[i], bhat=o$coef[2,1], se=o$coef[2,2], pval=o$coef[2,4], n=o$df[1] + o$df[2])
    }) %>% bind_rows()
  }) %>% bind_rows()
}

```



# Obtaining a count of proteins per ancestry with fdr less than 0.05 as either true or false 
```{r}
rcovs_2 <- phewas_covs_2(prot4, prospective_dat)
rcovs_2$fdr <- p.adjust(rcovs_2$pval, "fdr")
table(rcovs_2$fdr < 0.05, rcovs_2$pop)


```

#Extracting the proteins that are specific to Africans with fdr less than 0.05 

```{r}
# Extract proteins for a specific ancestry with FDR < 0.05
significant_proteins_ancestry_AFR <- rcovs_2$prot[rcovs_2$fdr < 0.05 & rcovs_2$pop == "AFR"]

# Display the significant proteins for the ancestry
print(significant_proteins_ancestry_AFR)

```

# Extract all proteins that are significantly associated with FDR < 0.05 per ancestry 
```{r}
# Extract significant proteins for all ancestries with FDR < 0.05
significant_proteins_all_ancestries_2 <- rcovs_2 %>%
  filter(fdr < 0.05) %>%
  select(pop, prot)

# Display the table of significant proteins for all ancestries
print(significant_proteins_all_ancestries_2)

```

# Obtaining a list of proteins that are shared by more than 1 ancestry 
```{r}
# Set the minimum number of ancestries for proteins to be considered shared
min_ancestry_count <- 4  # Change this number based on your criteria

# Count the total number of ancestries
total_ancestry_count <- n_distinct(significant_proteins_all_ancestries_2$pop)

# Group by protein and count the number of unique ancestries for each protein
proteins_shared_across_ancestries <- significant_proteins_all_ancestries_2 %>%
  group_by(prot) %>%
  summarise(
    unique_ancestry_count = n_distinct(pop),
    unique_ancestries = paste(unique(pop), collapse = ", ")
  )

# Filter proteins that are significant in more than one ancestry
shared_proteins <- proteins_shared_across_ancestries %>%
  filter(unique_ancestry_count >= min_ancestry_count)

# Display the table of shared proteins and their associated ancestries
print(shared_proteins)

```


```{r}
library(ggplot2)

# Plot shared proteins and their associated ancestries
ggplot(shared_proteins, aes(x = unique_ancestry_count, y = prot, fill = unique_ancestries)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Shared Proteins Across Ancestries",
    x = "Count",
    y = "Proteins"
  ) +
  theme_minimal()


```